---
title: "discussion_week7"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(sf)
library(terra)
library(tmap)
library(patchwork)
```

# Practice function making with rasters

Climate envelope models map where a species currently lives based on its climate conditions, then project where those conditions will exist under future climate scenarios. The key concern is whether suitable climate conditions will still exist anywhere or if a species’ climate envelope will shrink entirely. This threat is particularly acute for endemic species that are restricted to narrow geographic ranges.

Two California examples illustrate different vulnerabilities:

-   Yellow-billed magpie(Pica nutalli): Audubon projects climate change will eliminate 80% of its summer range and 100% of its winter range by 2080, potentially leaving no suitable habitat (Source)
-   Tule elk (Cervus canadensis nannodes): An arid-system ungulate with behavioral adaptability, with strong evidence of altering movement patterns during extreme drought; their range has become warmer and drier during the 20th century, but population has grown steadily (Source)

These simple species distribution models help identify the climate space each species occupies and the risk of losing it as climate conditions shift.This week, you will practice function making with rasters and create a generalizable workflow to determine the current climate envelope of the two species.

```{r}
magpie <- read_csv(here::here("data", "week7-data", "magpie_obvs.csv"))
tule_elk <- read_csv(here::here("data", "week7-data", "tule_elk_obvs.csv"))

bioclim_dir <- here::here("data","week7-data", "climate", "wc2.1_2.5m")
bioclim <- list.files(bioclim_dir, pattern = glob2rx("*.tif$"), 
                      full.names = TRUE)
bioclim_sort <- bioclim[
  # Sort filepaths based on numeric suffix
  order(
    # Extract numeric suffix of filenames and convert to numeric
    as.numeric(gsub(".*_(\\d+)\\.tif$", "\\1", bioclim))
    )
  ]
bioclim_rast <- rast(bioclim_sort)
```

## 2. Name raster layers

Update the `bioclim_rast` layer names to match the Bioclimatic variables using `names()`

```{r}
variables <- c("annualMeanTemp", "meanDiurnalRange", "isothermality", "tempSeasonality", "maxTempWarmMonth", "maxTempColdMonth", "tempAnnualRange", "meanTempWetQ", "meanTempDryQ", "meanTempWarmQ", "meanTempColdQ", "annualPrecip", "precipWetMonth", "precipDryMonth", "precipSeasonality", "precipWetQ", "precipDryQ", "precipWarmQ", "precipColdQ")

names(bioclim_rast) <- variables
```

## 3. Find geographic extent of species occurrence and crop raster

-   Use `st_as_sf` to convert and find the geographic extent of species occurrence data `magpie` using `st_bbox()`.
-   Crop `bioclim_rast` to match the extent of species occurrence data magpie.

```{r}

magpie_sf <- magpie %>% 
  rename(long = longitude,
         lat = latitude) %>%  
  drop_na(long, lat) %>%  
  st_as_sf(coords = c("long", "lat"), crs = 4326)

magpie_bbox <- st_bbox(magpie_sf)

bioclim_crop <- crop(bioclim_rast, magpie_bbox)
```

## 4. Extract points from species occurrence

Extract points from the cropped `bioclim_rast` for each occurrence in magpie.

```{r}
bioclim_pts <- as_tibble(extract(bioclim_crop, magpie_sf))
```

## 5. Create background values and extract points

Create “background values” by generating random sample points from the cropped `bioclim_rast`. Extract points from cropped `bioclim_rast` for the background values.

```{r}
# Create random sample points
background_values <- spatSample(bioclim_crop[["annualMeanTemp"]], 
                                size = nrow(magpie)*2, 
                                na.rm = TRUE,
                                as.points = TRUE) %>% 
  st_as_sf()


# Extract points from raster for background values
bioclim_random_pts <- as_tibble(extract(bioclim_crop, background_values))
```


## 6. Plot species climate niche

Plot species climate niche and background climate (temperature vs. precipitation)

```{r}
map_temp <- tm_shape(bioclim_crop[['annualMeanTemp']]) +
  tm_raster() +
tm_shape(magpie_sf) +
  tm_dots() 

map_precip <- tm_shape(bioclim_crop[['annualPrecip']]) +
  tm_raster()  +
tm_shape(magpie_sf)

tmap_arrange(map_temp, map_precip)
```

```{r}
plot_niche <- ggplot(data = bioclim_pts, aes(x = annualPrecip, y = annualMeanTemp)) +
  geom_point(shape = 16, color = "#3a5a40") +
  labs(x = "Annual precipitation",
       y = "Annual mean temperature", 
       title = "Species climate niche") +
  theme_bw()

plot_climate <- ggplot(data = bioClim_random_pts, aes(x = annualPrecip, y = annualMeanTemp)) +
  geom_point(shape = 16) +
  labs(x = "Annual precipitation",
       y = element_blank(), 
       title = "Background climate") +
  theme_bw()

plot_niche + plot_climate
```

## 7. Create generalizable workflow and apply

Modify previous steps into a generalizable workflow for other species occurrence data Try your new function with species occurrence data tule_elk.

```{r}
```


```{r}
geom_point(shape = 16, color = "#3a5a40") +
  labs(x = "Annual precipitation",
       y = "Annual mean temperature", 
       title = "Species climate niche") +
  theme_bw()

plot_climate <- ggplot(data = bioClim_random_pts, aes(x = annualPrecip, y = annualMeanTemp)) +
  geom_point(shape = 16) +
  labs(x = "Annual precipitation",
       y = element_blank(), 
       title = "Background climate") +
  theme_bw()

plot_niche + plot_climate
```

## 7. Create generalizable workflow and apply

Modify previous steps into a generalizable workflow for other species occurrence data Try your new function with species occurrence data tule_elk.

```{r}
climate_envelope <- function(clim_rast, clim_var1, clim_var2, occurences, species_name) {
  species_name_lower <- species_name |> 
    str_to_lower() |> 
    str_replace_all(" ", "_")
  
  occurences_sf <- occurences |> 
    rename(long = longitude,
           lat = latitude) |> 
    drop_na(long) |> 
    st_as_sf(coords = c("long", "lat"), crs = 4326)
  
  occurences_bbox <- st_bbox(occurences_sf)
  
  clim_crop <- crop(clim_rast, occurences_bbox)
  
  clim_pts <- as_tibble(extract(clim_crop, occurences_sf))
  
  random_pts <- spatSample(clim_rast[[clim_var1]],
                           na.rm = TRUE,
                           size = (nrow(occurences) * 2),
                           as.points = TRUE) |>
    st_as_sf()
  
  clim_random_pts <- as_tibble(terra::extract(clim_crop, random_pts))
  
  map_1 <- tm_shape(clim_crop[[clim_var1]]) +
    tm_raster(col.scale = tm_scale_continuous(values = "Blues")) +
    tm_shape(occurences_sf) +
    tm_dots(fill = "#3a5a40", size = 0.15) +
    tm_layout(legend.position = tm_pos_in("center", "bottom"),
              legend.orientation = "landscape",
              legend.bg.color = "white")
  
  map_2 <- tm_shape(clim_crop[[clim_var2]]) +
    tm_raster(col.scale = tm_scale_continuous(values = "-RdYlBu")) +
    tm_shape(occurences_sf) +
    tm_dots(fill = "#3a5a40", size = 0.15) +
    tm_layout(legend.position = tm_pos_in("center", "bottom"),
              legend.orientation = "landscape",
              legend.bg.color = "white")
  
  plot_1 <- ggplot(data = clim_pts, aes_string(x = clim_var1, y = clim_var2)) +
    geom_point(shape = 16, color = "#3a5a40") +
    labs(title = paste("Species climate niche of", str_to_sentence(species_name))) +
    theme_bw()
  
  plot_2 <- ggplot(data = clim_random_pts, aes_string(x = clim_var1, y = clim_var2)) +
    geom_point(shape = 16) +
    labs(y = element_blank(), 
         title = paste("Background climate of", str_to_sentence(species_name))) +
    theme_bw()
  
  assign(paste0(species_name_lower, "_map_1"), map_1, envir = .GlobalEnv)
  assign(paste0(species_name_lower, "_map_2"), map_2, envir = .GlobalEnv)
  assign(paste0(species_name_lower, "_plot_1"), plot_1, envir = .GlobalEnv)
  assign(paste0(species_name_lower, "_plot_2"), plot_2, envir = .GlobalEnv)
}
```

```{r}
climate_envelope(clim_rast = bioclim_rast,
                 clim_var1 = "annualPrecip",
                 clim_var2 = "annualMeanTemp",
                 occurences = tule_elk, 
                 species_name = "Tule Elk")

tule_elk_map_1
```

